---
title: "C3 R Bootcamp Bonus Day"
author: "Natalie Elphick"
date: "November 9th"
output:
  revealjs::revealjs_presentation:
    self_contained: false
    reveal_plugins: ["menu"]
    css: style.css
---

```{r, setup, include=FALSE}
library(tidyverse)
```

## 

<center>*Press the ? key for tips on navigating these slides*</center>

## Topics

-   RMD notebooks, R projects, and R environments

-   `apply` and `map` functions

-   Intro to Linear Models

-   How to find bioinformatics packages for your use case

-   ChatGPT Tips for Bioinformatics

-   Additional Resources

# RMD notebooks, R projects, and R environments

## R Markdown

-   A file format combining `R` code with [Markdown](https://www.markdownguide.org/basic-syntax/) for text formatting.
-   Designed for creating reproducible research reports in various formats (HTML, PDF, Word).
-   Let's create an `Rmd` file in `RStudio` to explore the basics of how they work:
-   File -\> New File -\> R Markdown

## R Markdown Advanced Usage

-   **Presentations:** Creating slides (like these) with [revealjs](https://github.com/rstudio/revealjs).
-   **Publications:** Authoring online books that combine narrative, code, and output with [bookdown](https://bookdown.org/).
-   **Interactive Documents:** Developing interactive tutorials or dashboards with [learnR](https://rstudio.github.io/learnr/) and other embedded applications.

## R Projects

-   A project is just a working directory designated with a `.RProj`
-   Opening the `.RProj` file sets the working directory to the location of the `.RProj` file
-   Helps implement a common best practice for data analysis
    -   Having the input data, R scripts, results and figures all under one directory
-   See [this chapter](https://r4ds.had.co.nz/workflow-projects.html#workflow-projects) for more information

## R Environments

-   At a basic level the environment is the data structure used to **scope** variables and functions.
-   **Scoping:** the concept of defining the visibility and accessibility of variables/functions within a program.
-   Similar to a named `list`, except:
    -   Every name must be unique.
    -   The names in an environment are not ordered.

# Apply and Map Functions

## apply()

-   Apply a function over a margin (1 = rows, 2 = columns) and return the results in an array

-   Alternative to for loops

```{r}
apply(X = cars, MARGIN = 2, FUN =  mean)
```

-   `apply()` works for matrices and data frames, there is also:

    -   `lapply()` for lists

    -   `vapply()` for vectors

## apply() : custom functions

-   You can supply a custom function within apply
-   These nameless functions are known as anonymous functions or lambda expressions

```{r}
apply(X = cars, MARGIN = 2,
      FUN =  function(col) {
        col_mean <- sum(col) / length(col)
        return(col_mean)
})
```

-   The above is much faster and more memory efficient than looping

## apply() : Converting for loops to apply

-   Since converting syntax can be tedious this is a good use case for ChatGPT
-   open the file

## purr::map() : tidyverse apply

-   [map()](https://purrr.tidyverse.org/reference/map.html) functions are used to apply a function to each element of a list or vector
-   They are type-stable, meaning they always return the same type of output (unlike apply)
-   `map_lgl()`, `map_int()`, `map_dbl()` and `map_chr()` return a vector of the indicated type

```{r}
library(purrr)
map(cars, ~ mean(.x)) 
```

## purr::map() (cont.)

-   Chaining multiple map calls can be a powerful tool to get complex outputs from a few lines of code

```{r, message=FALSE}
library(magrittr)
mtcars %>%
  split(mtcars$cyl) %>% 
  map(.f = function(df) lm(mpg ~ wt, data = df)) %>%
  map(summary) %>%
  map_dbl("r.squared")
```

# Intro to Linear Models in R

## Common Applications

-   RNA-Seq differential expression analyses
-   GWAS (for continuous traits)
-   eQTL analyses
-   Microarray data analyses

## Linear Model Background

$$
Y \sim \beta_0 + x_1\beta_1 + x_2\beta_2 ...x_n\beta_n + \epsilon
$$

-   $Y$ is our outcome

-   The covariates $x$ can be

    -   Continuous (age, weight, temperature, etc.)

    -   Categorical (treatment, batch, genotype, etc.)

-   The $\beta$'s are unknown parameters to be estimated.

-   The error term $\epsilon$ is assumed to be normally distributed with a variance that is constant across the range of the data.

## lm() : R Function for Linear Models

-   `lm()` expects a model formula
-   `lm()` does more than just linear regression
-   The formula is specified in the following format:

```         
Y ~ X
gene ~ age               # Regression
gene ~ sex               # One-way ANOVA
gene ~ age + genotype    # Analysis of covariance
```

-   See [Formulas in R](https://www.ucd.ie/ecomodel/Resources/R_formulae_WebVersion.html) for an in depth explanation of using formulas to specify models in R

## Example Dataset

```{r}
dat <- read.csv("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2018-September-Bioinformatics-Prerequisites/master/friday/lm_example_data.csv")
head(dat) %>%
  kableExtra::kable()
```

## Fitting the Model

```{r}
oneway_model <- lm(expression ~ treatment, data = dat)

summary(oneway_model)
```

## Understanding the Model Outputs

-   "Coefficients" refer to the $\beta$'s
-   "Estimate" is the estimate of each coefficient (effect size)
-   "Std. Error" is the standard error of the estimate
-   "t value" is the coefficient divided by its standard error
-   "Pr(\>\|t\|)"is the p-value for the coefficient

## Understanding the Model Outputs (cont.)

-   The residual standard error is the estimate of the variance of ϵ
-   Degrees of freedom is the sample size minus \# of coefficients estimated
-   R-squared is (roughly) the proportion of variance in the outcome explained by the model
-   The F-statistic compares the fit of the model as a whole to the null model (with no covariates)

## What do the Model Coefficients Mean?

-   R uses reference group coding or "treatment contrasts".
-   For categorical covariates, the first level alphabetically (or first factor level) is treated as the reference group.
-   The reference group is represented by the intercept.
-   Using control as a reference level is a common practice.

## Coefficients in the Example

-   Coefficients for other groups are the difference from the reference:

    (Intercept) = the mean for treatmentA\
    treatmentB = the mean for treatmentB - the mean for treatmentA\
    treatmentC = the mean for treatmentC - the mean for treatmentA\
    etc.

## broom::tidy : Summary Table of Model Fit

```{r}
broom::tidy(oneway_model) %>%
  arrange(desc(p.value)) %>%
  kableExtra::kable()
```

## Model Matrices

-   Tools like `limma` and `edgeR` use *design matrices* to specify the model

-   The design matrix $X$

    -   One row for each observation (sample)

    -   One column for each model coefficient (treatment)

```{r}
X <- model.matrix(~ treatment, data = dat)
head(X,4)
```

## Understanding Model Matrices

-   Linear regression case:

```{r}
X <- model.matrix(expression ~ temperature, data = dat)
X %>%
  head() %>%
  kableExtra::kable()
```

## Understanding Model Matrices (cont.)

-   One-way ANOVA:

```{r}
X <- model.matrix(expression ~ treatment, data = dat)
X %>%
  as.data.frame() %>%
  arrange(desc(treatmentB)) 
```

## Further Learning

-   Follow the [full tutorial](https://ucdavis-bioinformatics-training.github.io/2019-March-Bioinformatics-Prerequisites/thursday/linear_models.html) and complete the exercises
-   Learn about more advanced implementations of linear models:\
-   [Generalized Linear Models](https://bookdown.org/steve_midway/DAR/glms-generalized-linear-models.html)\
-   [Random Effects](https://bookdown.org/steve_midway/DAR/random-effects.html)

# ChatGPT Tips for Bioinformatics

## General Tips

-   Always confirm ChatGPT's outputs are correct
-   Provide as much detail as possible about the problem
-   Use separate chats for separate tasks/projects
-   Try the 'Custom Instructions' function that adds additional information to every query you run
-   The new release can visit webpages (GPT 4 only), which can help get more specific answers

## Code Tips

-   Commented R code yields better responses in my experience
-   Provide the code and error message in the same query
-   ChatGPT works well to convert syntax and improve your code:
    -   "Turn this loop into a function : [your code]"
    -   "Is there a better way to do this : [your code]"
-   Check out the file: `example_code/1_for_loop_to_apply.R` for an example use case

# Finding Bioinformatics Packages

## Key Questions

-   What assay was the package designed for?
-   When was the last release?
-   Is it maintained (frequent updates)?
-   Does it work on all operating systems?
-   Are other people using it?
-   Do they respond to github issues?
-   Is there a benchmarking paper?

## BioConductor and CRAN

-   Both of these have stringent requirements for packages they host (eg. for BioConductor they have to run on all major operating systems)

-   Prefer BioConductor packages if available over CRAN

-   Prefer CRAN packages over ones only hosted on GitHub

## Start with the Assay

-   Click [here](https://www.bioconductor.org/packages/release/BiocViews.html#___Sequencing) to go to BioC views
-   Pick the Assay you want to analyse
-   Pick the type of analysis you want to do
-   Find a package that does it
-   Find benchmarking papers to narrow the list of packages down
-   Find the vignette in on the package page

## Using Packages Outside of Intended Usage

-   Do NOT do this
-   If you absolutely have to, make sure you can justify this to reviewers before you waste your time



# Additional Resources

## R

-   [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/how-to-read-this-book.html) : Excellent R markdown reference

-   [R for Data Science](https://r4ds.hadley.nz/)

-   [ggplot2: elegant graphics for data analysis](https://ggplot2-book.org/)

-   [Advanced R](https://adv-r.hadley.nz/)

## Statistics

-   [Data Analysis in R](https://bookdown.org/steve_midway/DAR) : This book has more statistics details than *R for Data Science*
-   [Generalized Linear Models](https://bookdown.org/steve_midway/DAR/glms-generalized-linear-models.html)\
-   [Random Effects](https://bookdown.org/steve_midway/DAR/random-effects.html)

## RNA-seq Analysis

-   [RNA-seqlopedia](https://rnaseq.uoregon.edu/) : Everything you need to know about RNA-seq experiments
-   [RNA-seq Expression Units](https://luisvalesilva.com/datasimple/rna-seq_units.html) : Blog post on understanding common units
-   [Introduction to Single-Cell Analysis with Bioconductor](https://bioconductor.org/books/3.17/OSCA.intro/index.html) : Covers the basics of scRNA-seq analysis in R

## Dimensional Reduction

-   [Tutorial on PCA](https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/pca/) : PCA explained with R code examples
-   [Understanding UMAP](https://pair-code.github.io/understanding-umap/) : Short explanation with great visualizations, mainly useful for scRNA-seq analysis



# References

-   Linear models section was adapted from:
    -   [this page](https://ucdavis-bioinformatics-training.github.io/2019-March-Bioinformatics-Prerequisites/thursday/linear_models.html) created by the UC Davis Bioinformatics Core
